---
title: "BMEG424_Final Project"
output: html_document
date: "2025-04-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
We chose the study “Experimental Evolution Reveals Unifying Systems-Level Adaptations but Diversity Driving Genotypes” to investigate for our course project. The study investigates the evolution of six strains of Escherichia Coli subjected to adaptive laboratory evolution (ALE). The aim of this study was to characterize genotype-fitness maps of evolution for system-level properties, including metabolic and transcriptional regulatory networks. The authors took a multi-omic approach by studying the correlation between gene expression, gene mutation and metabolic flux. The goal was to uncover systems-level adaptations and identify consistent mutational patterns across a diverse genetic background. 

Our reanalysis will focus on processing the raw sequencing data from this study to confirm the reproducibility and robustness of the authors’ methodology. We chose this specific study due to its comprehensive coverage of core concepts in genomic data analysis, as well as the large amount of publicly available sequencing data. WE WILL IMPROVE AND LOOK FOR OVERSIGHTS. Our analysis will include quality control, genome indexing and paired end alignment using automated pipelines and workflows. 

Potential oversights:

No automated pipeline?
No mention of quality control?
No report of alignment stats?

## Quality Control for Fastq Files

## Alignment 

The paper aligned all reads to their respective genome using Bowtie v1.1.2 (Langmead et al., 2009) with the following flags: -m 1 --best --strata." Bowtie is an older, ultra-fast aligner for short reads, however it is not splice aware and doesn't allow for gaps (no insertions/deletions). The flags that were used discards reads that map to more than one location and ensure the best alignment is selected. This setup appears to be chosen based on emphasizing specificity versus sensitivity. 

We chose to use BWA, which is a more modern aligner which offers better handling of mismatches and works well with longer reads. We attempted to simulate their results using this aligner by filtering out uniquely mapped reads. 

is the data longer or noisier gotta check

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
reference_map = {
    "SRR12170043": "NC_007779.1",
    "SRR12170044": "NC_007779.1",
    "SRR12170045": "NC_007779.1",
    "SRR12170046": "NC_007779.1",
    "SRR12170047": "NC_007779.1",
    "SRR12170048": "NC_007779.1",
    "SRR12170049": "NC_007779.1",
    "SRR12170050": "NC_000913.3",
    "SRR12170051": "NC_000913.3",
    "SRR12170052": "NC_000913.3",
    "SRR12170053": "NC_012971.2",
    "SRR12170054": "NC_000913.3",
    "SRR12170055": "NC_000913.3",
    "SRR12170056": "NC_000913.3",
    "SRR12170057": "NC_007779.1",
    "SRR12170058": "NC_000913.3",
    "SRR12170059": "NC_000913.3",
    "SRR12170060": "NC_000913.3",
    "SRR12170061": "NC_000913.3",
    "SRR12170062": "NC_000913.3",
    "SRR12170063": "NC_010468.1",
    "SRR12170064": "NC_012971.2",
    "SRR12170065": "NC_010468.1",
    "SRR12170066": "NC_010468.1",
    "SRR12170071": "NC_007779.1",
    "SRR12170072": "NC_007779.1",
    "SRR12170075": "NC_012971.2",
    "SRR12170076": "NC_012971.2"
}

rule all:
    input:
        expand("aligned/{sample}.unique.sorted.bam", sample=reference_map.keys())

rule download_sra:
    output:
        "sra/{sample}.sra"
    shell:
        "prefetch {wildcards.sample} -O sra"

rule sra_to_fastq:
    input:
        "sra/{sample}.sra"
    output:
        r1="data/{sample}_1.fastq.gz",
        r2="data/{sample}_2.fastq.gz"
    shell:
        "fastq-dump --split-files --gzip --outdir data {input}"

rule index_ref:
    input:
        "ref_genomes/{ref}.fna"
    output:
        "ref_genomes/{ref}.fna.bwt"
    shell:
        "bwa index {input}"

rule align:
    input:
        r1="data/{sample}_1.fastq.gz",
        r2="data/{sample}_2.fastq.gz",
        ref=lambda wildcards: f"ref_genomes/{reference_map[wildcards.sample]}.fna",
        idx=lambda wildcards: f"ref_genomes/{reference_map[wildcards.sample]}.fna.bwt"
    output:
        "aligned/{sample}.sam"
    shell:
        "bwa mem {input.ref} {input.r1} {input.r2} > {output}"

rule sam_to_bam:
    input:
        "aligned/{sample}.sam"
    output:
        "aligned/{sample}.bam"
    shell:
        "samtools view -bS {input} > {output}"

rule filter_unique:
    input:
        "aligned/{sample}.bam"
    output:
        "aligned/{sample}.unique.bam"
    shell:
        "samtools view -b -q 1 {input} > {output}"

rule sort_bam:
    input:
        "aligned/{sample}.unique.bam"
    output:
        "aligned/{sample}.unique.sorted.bam"
    shell:
        "samtools sort -o {output} {input}"

```
